Certified Knowledge Compilation \\ with Application to Formally Verified Model Counting

### 2024-01-05 doc/JAIR/jair-paper.tex
Minor polishing
Fix up Extensions section

As with other representations, each node $\nodeu$ in an SDD denotes a
Boolean formula $\phi_{\nodeu}$.  Leaf nodes correspond to $\tautology$, $\nil$, and literals.
Each nonterminal node has $2k$ children,
organized as pairs $\nodep_i$ and $\nodes_i$ for $1\leq i \leq k$.
The dependency set for the node must be partitioned into disjoint sets $X_p$ and $X_s$ such that
$\dependencyset(\phi_{\nodep_i}) \subseteq X_p$ and
$\dependencyset(\phi_{\nodes_i}) \subseteq X_s$ for all $1 \leq i \leq k$.
Children $\nodep_1, \nodep_2, \ldots, \nodep_k$ serve as a set of
\emph{selector} functions with mutually disjoint models.  That is,
$\modelset(\phi_{\nodep_i}) \cap \modelset(\phi_{\nodep_j}) =
\emptyset$ for $1 \leq i < j \leq k$.
The Boolean formula associated with node $\nodeu$ is then
\begin{eqnarray}
\phi_{\nodeu} & = & \bigvee_{1 \leq i \leq k} \left[\phi_{\nodep_i} \land \phi_{\nodes_i}\right] \label{eqn:sdd}
\end{eqnarray}

An SDD can be translated into a POG, where an SDD node with $2k$
children expands into $k$ product nodes and $k-1$ sum nodes.
In particular, assume that each nonterminal SDD node $\nodev$ can be represented by a
subgraph in the POG with root node $\hat{\nodev}$.  For leaf node $\nodev$, its counterpart $\hat{\nodev}$ is
the associated constant or literal, rather than a POG node.  Figure~\ref{fig:sdd:pog} then
illustrates how node $\nodeu$ with $2k$ children would be expanded
into a subgraph with root node $\hat{\nodeu}$.  We can see that each
of the product nodes has arguments with disjoint dependency sets, and
each of the sum nodes has disjoint sets of models.

Several different methods could be used to generate a CPOG
certification from an SDD\@.  For the forward implication proof, the
monolithic approach we have used could generate the proof via a single
call to a proof-generating SAT solver.  For an SDD generated by the
bottom up approach originally proposed by \citeA{darwiche:ijcai:2011},
the generator could be modified to generate proof steps for each
recursive call during the construction, as is done to generate proofs
of unsatisfiability with proof-generating, BDD-based SAT
solvers~\cite{Jussila:2006,ebddres,bryant:fmcad:2022,bryant:tocl:2023}.
For an SDD generated by a top-down process, a structural method
similar to the one we have described could be applied.  A further
requirement would be to generate mutual exclusion proofs for the $k-1$
sum nodes illustrated in Figure~\ref{fig:sdd:pog}.  These would be
based on the disjoint model property of the $k$ nodes of the form
$\hat{\nodep}_i$.  One approach would be to simply make $k-1$ calls a
proof-generating SAT solver.

@@

Kimmig's formulation of algebraic model
counting~\cite{kimmig:jal:2017} is more general than ours.  It allows
the algebraic structure to be a \emph{semiring}.  A commutative
semiring $\semiring$ obeys all properties of a commutative ring, except that the
elements of the set need not have additive inverses.
We can then define the {\em semiring evaluation problem} as computing
  \begin{equation}
    \begin{array}{rcl}
    \srep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:srep}
    \end{array}
  \end{equation}
  where sum $\sum$ is computed according to the semiring addition operation $\radd$ and product $\prod$
  is computed according to the semiring product operation $\rmul$.

@@

Semiring evaluation can be performed via knowledge compilation
by requiring 
that the representation generated by
the compiler obeys a property known as
\emph{smoothness}~\cite{darwiche:jair:2002,shih:nips:2019}.  Within our formulation,
a partitioned-operation formula is smooth when all arguments to each
sum operation have identical dependency sets.  That is, every sum
operation $\bigvee_{1 \leq i \leq k} \phi_i$ has
$\dependencyset(\phi_i) = \dependencyset(\phi_j)$ for $1 \leq i < j
\leq k$.  Smoothness can be ensured by adding redundant formulas to
artificially introduce variables.  For example, if subformula $\phi_i$
lacks having variable $x$ in its dependency set, it can be replaced by
$(x \por \obar{x}) \pand \phi_i$.  When a knowledge compiler generates
a representation that is both in normal form and is smooth, then a semiring evaluation of the formula
can proceed by first assigning each literal $\lit$ the value $w(\lit)$.
Then the product and sum operations are evaluated in manners analogous to
(\ref{eqn:ring:product}) and
(\ref{eqn:ring:sum}).

Our POG representation can support evaluation of semiring formulas by
imposing the restriction that the POG is in normal form and that is smooth.
Given a smoothed
dec-DNNF graph generated by a knowledge compiler, our toolchain will
convert this into a smooth POG in normal form and verify its
equivalence to the input formula.  Full verification would also require checking that the POG is smooth.

>>>>

As with other representations, each node $\nodeu$ in an SDD denotes a
Boolean formula $\phi_{\nodeu}$.  Leaf nodes correspond to Boolean constants and literals.
Each nonterminal node has $2k$ children,
organized as pairs $\nodep_i$ and $\nodes_i$ for $1\leq i \leq k$.
The dependency set for the node must be partitioned into disjoint sets $X_p$ and $X_s$ such that
$\dependencyset(\phi_{\nodep_i}) \subseteq X_p$ and
$\dependencyset(\phi_{\nodes_i}) \subseteq X_s$ for all $1 \leq i \leq k$.
Children $\nodep_1, \nodep_2, \ldots, \nodep_k$ serve as a set of
\emph{selector} functions with mutually disjoint models.  That is,
$\modelset(\phi_{\nodep_i}) \cap \modelset(\phi_{\nodep_j}) =
\emptyset$ for $1 \leq i < j \leq k$.
The Boolean formula associated with node $\nodeu$ is then
\begin{eqnarray}
\phi_{\nodeu} & = & \bigvee_{1 \leq i \leq k} \left[\phi_{\nodep_i} \land \phi_{\nodes_i}\right] \label{eqn:sdd}
\end{eqnarray}

An SDD can be translated into a POG, where an SDD node with $2k$
children expands into $k$ product nodes and $k-1$ sum nodes.
In particular, assume that each nonterminal SDD node $\nodev$ can be represented by a
subgraph in the POG with root node $\hat{\nodev}$.  For leaf node $\nodev$, its counterpart $\hat{\nodev}$ is
the associated terminal node.  Figure~\ref{fig:sdd:pog} then
illustrates how node $\nodeu$ with $2k$ children would be expanded
into a subgraph with root node $\hat{\nodeu}$.  We can see that each
of the product nodes has arguments with disjoint dependency sets, and
each of the sum nodes has disjoint sets of models.

Several different methods could be used to generate a CPOG
certification from an SDD\@.  For the forward implication proof, the
monolithic approach we have used could generate the proof via a single
call to a proof-generating SAT solver.  For an SDD generated by the
bottom up approach originally proposed by \citeA{darwiche:ijcai:2011},
the generator could be modified to generate proof steps for each
recursive call during the construction, as is done to generate proofs
of unsatisfiability with proof-generating, BDD-based SAT
solvers~\cite{Jussila:2006,ebddres,bryant:fmcad:2022,bryant:tocl:2023}.
For an SDD generated by a top-down process, a structural method
similar to the one we have described could be applied.  A further
requirement would be to generate mutual exclusion proofs for the $k-1$
sum nodes illustrated in Figure~\ref{fig:sdd:pog}.  These would be
based on the disjoint model property of the $k$ nodes of the form
$\hat{\nodep}_i$.  One approach would be to simply make $k-1$ calls to a
proof-generating SAT solver.

@@

The formulation of algebraic model
counting by \citeA{kimmig:jal:2017} is more general than ours.  It allows
the algebraic structure to be a \emph{semiring}.  A commutative
semiring $\semiring$ obeys all properties of a commutative ring, except that the
elements of the set need not have additive inverses.
We can then define the {\em semiring evaluation problem} as computing
  \begin{equation}
    \begin{array}{rcl}
    \srep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:srep}
    \end{array}
  \end{equation}
  where sum $\sum$ is computed according to the semiring addition operation $\radd$ and product $\prod$
  is computed according to the semiring product operation $\rmul$.

@@

Semiring evaluation can be performed via knowledge compilation
by requiring 
that the representation generated by
the compiler obeys a property known as
\emph{smoothness}~\cite{darwiche:jair:2002,shih:nips:2019}.  Within our formulation,
a partitioned-operation formula is smooth when all arguments to each
sum operation have identical dependency sets.  That is, every sum
operation $\bigvee_{1 \leq i \leq k} \phi_i$ has
$\dependencyset(\phi_i) = \dependencyset(\phi_j)$ for $1 \leq i < j
\leq k$.  Smoothness can be ensured by adding redundant formulas to
artificially introduce variables.  For example, if subformula $\phi_i$
lacks having variable $x$ in its dependency set, it can be replaced by
$(x \por \obar{x}) \pand \phi_i$.  When a knowledge compiler generates
a representation in negation normal form that is smooth, then a semiring evaluation of the formula
can proceed by first assigning each literal $\lit$ the value $w(\lit)$.
Then the product and sum operations are evaluated in manners analogous to
(\ref{eqn:ring:product}) and
(\ref{eqn:ring:sum}).

Our POG representation can support evaluation of semiring formulas by
imposing the restriction that the POG is in negation normal form and that it is smooth.
Given a smoothed
dec-DNNF graph generated by a knowledge compiler, our toolchain will
convert this into a smooth POG in normal form and verify its
equivalence to the input formula.  Of course, full verification would also require checking that the POG is smooth.
